Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-20 introduction.tex
more changes to intro, overview

Many phenomena in astrophysics and cosmology require a high spatial and temporal dynamical range and there are a number of numerical techniques which have been used to model such problems.  At the current time, the most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful, particularly for those situations dominated by gravity.  Its development, however, is still at a relatively early stage when compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{, laney-1998, toro-1997, Woodward84}.

>>>>

Many phenomena in astrophysics and cosmology require a high spatial and temporal dynamical range and there are a number of numerical techniques which have been used to model such problems.  At the current time, the most common is a gridless, fully Lagrangian method call Smoothed Particle Hydrodynamcs \citep[SPH;][]{Lucy77, SPH} which has been very successful, particularly for those situations dominated by gravity.  Its development, however, is still at a relatively early stage when compared to the effort put into Eulerian, grid-based hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997, Woodward84}.

### 2013-03-20 numerical-radiation-fld.tex
more changes to intro, overview

Enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities, 
\begin{eqnarray}
  \label{eq:fld_radiation}
  \partial_t E_r + \frac1a \nabla\cdot\left(E_r \vec{v}_b\right) &=& 
    \nabla\cdot\left(D\nabla E_r\right) -
    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
  \label{eq:fld_heating}
  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
  \label{eq:fld_chemistry}
  \partial_t {\mathrm n}_i + 
    \frac1a \nabla\cdot\left({\mathrm n}_i \vec{v}_b\right) &=& 
    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
\end{eqnarray}
where $E_r$ is a grey radiation energy density, $e_c$ is the
internal energy correction due to photo-heating and chemical cooling,
and ${\mathrm n}_i$ correspond to the chemical number densities of
HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
number density.  Here, we define the grey radiation energy density
through first assuming a fixed frequency spectrum, i.e.
$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
defining the integrated quantity
\begin{equation}
\label{eq:grey_radiation_energy}
   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
\end{equation}
In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
opacity, $\eta$ is the field of radiation sources, $G$ provides the
radiation induced photo-heating, $\Lambda$ provides the chemical
cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_chemistry}) is solved
independently of Enzo's hydrodynamics, gravity and dark-matter solvers
(sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of equations (\ref{eq:fld_radiation}) and
(\ref{eq:fld_chemistry}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given time step to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

We then evolve the system (\ref{eq:fld_heating}-\ref{eq:fld_chemistry}):  
$(E_r^{n+1}, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1},
{\mathrm n}_i^{n+1})$.  Due to the lack of spatial derivatives (since
advection is handled elsewhere), this system is a coupled system of
nonlinear ordinary differential equations.  This utilizes an implicit
quasi-steady-state approximation, formulated as follows.  We consider
the modified equations,
\begin{eqnarray}
  \label{eq:fld_heating_qss}
  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c +
    G\left(\bar{E}_r,\bar{\mathrm n}_i\right) - 
    \Lambda\left(\bar{E}_r,\bar{\mathrm n}_i\right), \\
  \label{eq:fld_chemistry_qss}
  \partial_t {\mathrm n}_i &=& \alpha_{i,j}\left(\bar{e}_c\right)
    {\mathrm n}_e \bar{\mathrm n}_j - {\mathrm n}_i 
    \Gamma_i^{ph}\left(\bar{E}_r\right), \quad i=1,\ldots,N_s,
\end{eqnarray}
where we have defined the time-centered ``background'' states
$\bar{E}_r = \left(E_r^{n}+E_r^{n+1}\right)/2$, 
$\bar{\mathrm n}_i = \left({\mathrm n}_i^{n}+{\mathrm n}_i^{n+1}\right)/2$
and $\bar{e}_c = \left(e_c^{n}+e_c^{n+1}\right)/2$.  These equations
may each be solved {\em analytically} for their solution at the time
$t$, which we denote by
\begin{eqnarray}
  \label{eq:fld_heating_qss_sol}
  e_c(t) &=& \text{sol}_{e}\left(t,\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right), \\
  \label{eq:fld_chemistry_qss_sol}
  {\mathrm n}_i(t) &=& \text{sol}_{\mathrm n_i}
  \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right), \quad i=1,\ldots,N_s. 
\end{eqnarray}
We then define a nonlinear system of equations to compute the
time-evolved solutions $\left(E_r^{n+1}, e_c^{n+1}, 
{\mathrm n}_i^{n+1})\right)$ as
\begin{eqnarray}
  \label{eq:fld_heating_qss_fe}
  f_e(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& e_c^{n+1} -
    \text{sol}_{e}\left(t^{n+1},\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right)
    = 0, \\
  \label{eq:fld_chemistry_qss_fn}
  f_{\mathrm n_i}(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& 
    {\mathrm n}_i(t) - \text{sol}_{\mathrm n_i}
    \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right)=0, \quad
    i=1,\ldots,N_s.  
\end{eqnarray}
This system of $N_s+1$ nonlinear equations is solved using a damped
fixed point iteration, 
\[
   U_i = U_i - \lambda f_i(U), \quad i=1,\ldots,N_s+1,
\]
where $U$ is a vector containing the solutions to equations
(\ref{eq:fld_heating_qss_fe}-\ref{eq:fld_chemistry_qss_fn}).  In this
iteration, for the first 50 sweeps we use $\lambda=1$, and for more
challenging problems if that does not suffice we switch to a damping
parameter of $\lambda=0.1$.

>>>>

Enzo's field-based radiation solver focuses on
a flux-limited diffusion approximation for cosmological radiative
transfer, with couplings to both the gas energy and chemical number
densities.
%\begin{eqnarray}
%  \label{eq:fld_radiation}
%  \partial_t E_r + \frac1a \nabla\cdot\left(E_r \vec{v}_b\right) &=& 
%    \nabla\cdot\left(D\nabla E_r\right) -
%    \frac{\dot{a}}{a} E_r - c\kappa E_r + \eta, \\
%  \label{eq:fld_heating}
%  \partial_t e_c &=& -\frac{2\dot{a}}{a} e_c + G - \Lambda, \\
%  \label{eq:fld_chemistry}
%  \partial_t {\mathrm n}_i + 
%    \frac1a \nabla\cdot\left({\mathrm n}_i \vec{v}_b\right) &=& 
%    \alpha_{i,j} {\mathrm n}_e {\mathrm n}_j - {\mathrm n}_i
%    \Gamma_i^{ph}, \quad i=1,\ldots,N_s,
%\end{eqnarray}
%where $E_r$ is a grey radiation energy density, $e_c$ is the
%internal energy correction due to photo-heating and chemical cooling,
%and ${\mathrm n}_i$ correspond to the chemical number densities of
%HI, HII, HeI, HeII and HeIII, and ${\mathrm n}_e$ is the electron
%number density.  Here, we define the grey radiation energy density
%through first assuming a fixed frequency spectrum, i.e.
%$E_{\nu}(\nu,\vec{x},t) = \tilde{E}_r(\vec{x},t) \chi(\nu)$, and then
%defining the integrated quantity
%\begin{equation}
%\label{eq:grey_radiation_energy}
%   E_r(\vec{x},t) \equiv \int_{\nu_0}^{\infty}
%   E_{\nu}(\nu,\vec{x},t)\,\mathrm d\nu \  = \ 
%   \tilde{E}_r(\vec{x},t) \int_{\nu_0}^{\infty} \chi(\nu)\,\mathrm d\nu.
%\end{equation}
%In addition, $D$ in (\ref{eq:fld_radiation}) is the {\em Larsen
%square-root flux-limiter} (see \cite{Morel2000}), $\kappa$ is the total
%opacity, $\eta$ is the field of radiation sources, $G$ provides the
%radiation induced photo-heating, $\Lambda$ provides the chemical
%cooling, $\alpha_{i,j}$ are the reaction rate coefficients defining
%the interactions between species ${\mathrm n}_i$ and ${\mathrm n}_j$,
%and $\Gamma_i^{ph}$ are the radiation induced photo-ionization rates.

The system of equations
(\ref{eq:fld_radiation}-\ref{eq:fld_heating}) along with the chemical network 
(eq.~\ref{eq:species_evolution} is solved
independently of Enzo's hydrodynamics, gravity and dark-matter solvers
(sections \ref{sec.hydro.ppm}-\ref{sec.ov.nbody}), thereby allowing
the advective portions of equations (\ref{eq:fld_radiation}) and
(\ref{eq:fld_chemistry}) to be taken care of by the fluid solvers.
Due to the disparate time scales between radiation transport and chemical
ionization and heating, the remainder of these equations is solved
using an operator-split algorithm.  Within a given time step to evolve
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1}, {\mathrm
n}_i^{n+1})$, we first evolve equation (\ref{eq:fld_radiation}):
$(E_r^n, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n}, {\mathrm
  n}_i^{n})$.  This uses an implicit Euler time discretization, and a
second-order centered finite-difference spatial discretization,
resulting in a large linear system of equations.  These are solved
using a preconditioned conjugate gradient iteration, where the
preconditioner consists of a geometric multigrid solver.  Both of
these linear solvers are provided by the HYPRE linear solver library
(see \cite{FalgoutYang2002} and \cite{hypre-manual}).

We then evolve the system (\ref{eq:fld_heating}-\ref{eq:fld_chemistry}):  
$(E_r^{n+1}, e_c^n, {\mathrm n}_i^n) \to (E_r^{n+1}, e_c^{n+1},
{\mathrm n}_i^{n+1})$.  Due to the lack of spatial derivatives (since
advection is handled elsewhere), this system is a coupled system of
nonlinear ordinary differential equations.  This utilizes an implicit
quasi-steady-state approximation, formulated as follows.  We consider
the modified equations,
\begin{eqnarray}
  \label{eq:fld_heating_qss}
  \frac{\partial e_c}{\partial t} &=& -\frac{2\dot{a}}{a} e_c +
    \Gamma\left(\bar{E}_r,\bar{\mathrm n}_i\right) - 
    \Lambda\left(\bar{E}_r,\bar{\mathrm n}_i\right), \\
  \label{eq:fld_chemistry_qss}
  \frac{\partial {\mathrm n}_i}{\partial t} &=& k_{i,j}\left(\bar{e}_c\right)
    {\mathrm n}_e \bar{\mathrm n}_j - {\mathrm n}_i 
    \Gamma_i^{ph}\left(\bar{E}_r\right), \quad i=1,\ldots,N_s,
\end{eqnarray}
where we have defined the time-centered ``background'' states
$\bar{E}_r = \left(E_r^{n}+E_r^{n+1}\right)/2$, 
$\bar{\mathrm n}_i = \left({\mathrm n}_i^{n}+{\mathrm n}_i^{n+1}\right)/2$
and $\bar{e}_c = \left(e_c^{n}+e_c^{n+1}\right)/2$.  These equations
may each be solved {\em analytically} for their solution at the time
$t$, which we denote by
\begin{eqnarray}
  \label{eq:fld_heating_qss_sol}
  e_c(t) &=& \text{sol}_{e}\left(t,\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right), \\
  \label{eq:fld_chemistry_qss_sol}
  {\mathrm n}_i(t) &=& \text{sol}_{\mathrm n_i}
  \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right), \quad i=1,\ldots,N_s. 
\end{eqnarray}
We then define a nonlinear system of equations to compute the
time-evolved solutions $\left(E_r^{n+1}, e_c^{n+1}, 
{\mathrm n}_i^{n+1})\right)$ as
\begin{eqnarray}
  \label{eq:fld_heating_qss_fe}
  f_e(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& e_c^{n+1} -
    \text{sol}_{e}\left(t^{n+1},\bar{E}_r,\bar{\mathrm n}_i,e_c^n\right)
    = 0, \\
  \label{eq:fld_chemistry_qss_fn}
  f_{\mathrm n_i}(e_c^{n+1},{\mathrm n}_i^{n+1}) &\equiv& 
    {\mathrm n}_i(t) - \text{sol}_{\mathrm n_i}
    \left(t,\bar{E}_r,\bar{e_c},\mathrm n_i^n\right)=0, \quad
    i=1,\ldots,N_s.  
\end{eqnarray}
This system of $N_s+1$ nonlinear equations is solved using a damped
fixed point iteration, 
\[
   U_i = U_i - \lambda f_i(U), \quad i=1,\ldots,N_s+1,
\]
where $U$ is a vector containing the solutions to equations
(\ref{eq:fld_heating_qss_fe}-\ref{eq:fld_chemistry_qss_fn}).  In this
iteration, for the first 50 sweeps we use $\lambda=1$, and for more
challenging problems if that does not suffice we switch to a damping
parameter of $\lambda=0.1$.

### 2013-03-20 numerical-radiation-raytracing.tex
Adding description of photo-evaporation test and some minor changes to the Moray section.
more changes to intro, overview

Local radiation from stars and black holes strongly affects their
surroundings.  Radiation transport is a well-studied problem; however,
its treatment in multidimensional calculations is difficult because of
the dependence on seven variables -- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  The radiative transfer equation in comoving coordinates
\citep{Gnedin97} reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
specific intensity in units of energy per time $t$ per solid angle per
unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
ratio of scale factors at the current time and time of emission.  The
second term represents the propagation of radiation, where the factor
$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
both the cosmological redshift and dilution of radiation.  On the
right hand side, the first term considers the absorption coefficient
$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
that includes any point sources of radiation or diffuse radiation.

We can make some appropriate approximations to reduce the complexity
of Equation (\ref{eq:rteqn}) in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

>>>>

Local radiation from stars and black holes strongly affects their
surroundings.  Radiation transport is a well-studied problem; however,
its treatment in multidimensional calculations is difficult because of
the dependence on seven variables -- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  Here we briefly describe the ray tracing implementation \moray,
which is presented in full detail in \citet{Wise11_Moray} with seven
code tests and six applications.

%The radiative transfer equation in comoving coordinates
%\citep{Gnedin97} reads
%%
%\begin{equation}
%  \label{eq:rteqn}
%  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
%  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
%  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
%  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
%\end{equation}
%%
%Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
%specific intensity in units of energy per time $t$ per solid angle per
%unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
%constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
%ratio of scale factors at the current time and time of emission.  The
%second term represents the propagation of radiation, where the factor
%$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
%both the cosmological redshift and dilution of radiation.  On the
%right hand side, the first term considers the absorption coefficient
%$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
%$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
%that includes any point sources of radiation or diffuse radiation.
%GB: this was moved to the equations section.

We solve the radiative transfer equation in comoving coordinates (given by eq.~\ref{eq:rteqn}).
We can make some appropriate approximations to reduce the complexity
of this equation in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.
